Chemical Tagging Using Minimum Message Length. I.\\ Gaussian Mixture Modelling

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
describe the things we want to infer from chemical abundances

\usepackage{bm}

@@

@@

% Some sentence about using a transformative number of field stars to unravel
% the galaxy.
% Chemical fingerprint.

%The Milky Way is our best laboratory for understanding galaxy formation and
%evolution. 
Making strong inferences about the Milky Way's formation and evolution will
require a transformative number of stars with precise chemical abundances.
Wide-field multiplexed surveys are poised to deliver these data. 
The APO Galactic Evolution Experiment \citep[\apogee, e.g.,][]{apogee} and the 
Gaia-ESO survey \citep[\ges;][]{Gilmore_2012,Randich_2013} are already 
delivering $10^5$ stars with 20--40 precise abundances, and the Galactic
Archaeology with \hermes\ \citep[\galah;][]{DeSilva_2015} survey seeks to
derive $\approx$30 precise abundances for $10^6$ stars.
Future ground-based surveys will each derive abundances from $\approx$20
million spectra \citep[e.g., \weave, \fourmost;][]{weave,4most} in each
hemisphere, and the radial velocity spectrometer on \gaia\ can deliver a few
abundances for $\approx$150 million stars.

At this scale, chemical tagging represents a formidable parameter estimation
and model selection problem.

it represents a formidable parameter
estimation and model selection problem.
The scale of the problem is also uncertain.
A \emph{transformative} sample of stars with precise chemical abundance
measurements is presumably required to fully realise chemical tagging.
What can be learned from $10^4$ field stars with infinitely precise chemical
abundances? Can definitive inferences be made about the star formation history
of the Milky Way? Will $10^5$ field stars allow us to inform the initial mass
function? Are $10^6$ stars required before we can quantify the extent of
radial mixing and galactic-scale dynamics?

We explicitly list the components of the problem in detail before
reviewing progress in the literature.

Given a transformative sample of stars with precise chemical abundance
measurements, the number of star clusters that represent the observed sample is
unknown.

Progress to date has focussed on addressing individual aspects of this problem.

Some progress has been made towards chemical tagging in simplistic experiments, 
but the full extent of chemical tagging has yet to be realised. Indeed, in this 
\article\ we assert that even \emph{the full extent of the problem} of chemical 
tagging has yet to be realised. Before reviewing the existing literature on
what has been done with chemical tagging, we will outline the necessary
components to perform chemical tagging.

\subsection{What's needed to chemically tag a galaxy?}
\label{section:the-problem}

Given a set of stars that have been observed spectroscopically, and detailed
chemical abundances have been derived, the following issues will arise when
trying to perform chemical tagging:

\begin{enumerate}
\item The number of star clusters (or star-forming events) where at least one
      star was observed and detailed chemical abundances were derived.
\item The number of latent factors (e.g., supernovae yields, AGB yields).
      Within the chemical tagging literature, this is commonly referred to as
      the number of `effective chemical dimensions'.
\item The multivariate factor loads of individual latent factors. For example,
      if core-collapse supernovae and AGB stars had both contributed to the
      chemical abundances of all stars in the sample, then the
      \emph{factor loads} would be the \emph{yield} of the supernovae, or the
      yield of relative abundances that were produced by the AGB star. Each
      star will be affected differently by these yields (see next item).
\item The relative \emph{scoring} of different latent factors on each star.
      This is a scaling that is applied to the latent yields. Continuing our
      analogy on yields from core-collapse supernovae and AGB stars, the
      \emph{factor scores} can be thought of as the relative contributions of
      the different factor loads, for a single star.
\item The relative weighting (yet another latent factor, which is however not
      common between all mixtures) of individual mixtures.
\item The multivariate mean abundances and covariance matrices of each star
      cluster, where the covariance matrix includes off-diagonal terms that
      describe the \emph{data} correlations (e.g., the correlations that result
      from trying to measure multiple abundances from the same data).
\item The intrinsic \emph{specific variances} of the individual abundances.
\item Atomic diffusion, veiling, and thermohaline mixing: The surface chemical
      abundances of stars \emph{will change} over its lifetime.
\item Probabilistic priors (perhaps very hard priors) on joint responsibilities
      of stars that are specified from astrometric constraints.
\item There are missing data. 
\item There are \emph{upper limits} on some data.\footnote{
        Spectroscopists are probably the worst in the Universe when it comes to
        reporting upper limits. They will report an `\emph{upper limit}', when
        what might mean is the \emph{absolute limit} where there is zero
        likelihood that the true value is above the given point estimate (which might as well be $+\infinity$, or
        they may describe a \emph{pseudo-1$\sigma$} limit which is estimated
        from the locally (yet covariant!) heteroskedastic noise in the data (which often might as well be $\approx$0).
        In reality, an upper limit ought to be given as a probability density
        function that is skewed in one direction, perhaps similar to the shape
        of a log-normal distribution. In any case, the confidence to which they
        describe their `\emph{upper limit}' ought to be given.}
\item There are correlations in the measurements that result from trying to
      measure two (or more) abundances from the same data. As an example,
      consider a rare -- yet important -- chemical abundance that has a single
      atomic absorption line which is blended with a more common elemental
      abundance. In all likelihood, there will be a strong positive correlation
      between the abundance derived between the rare and more common elemental
      abundance.
\end{enumerate}

@@

>>>>

\usepackage{bm}
\usepackage{verbatim} % for \begin{comment}'ing out sections.

@@

% Common terms
\def\teff{T_{\rm eff}}
\def\logg{\log{g}}

@@

%Their signature of formation is locked up in their detailed chemical fingerprint

%The Milky Way is our best laboratory for understanding galaxy formation and
%evolution. 
Making strong inferences about the Milky Way's formation and evolution from
chemical tagging will require a transformative number of stars with precise
abundances.
Ground- and space-based surveys are poised to deliver those data this decade. 
The Apache Point Observatory Galactic Evolution Experiment 
\citep[\apogee, e.g.,][]{apogee} and the Gaia-ESO survey 
\citep[\ges;][]{Gilmore_2012,Randich_2013} are already delivering $10^5$ stars 
with 20--40 precise abundances, and the Galactic Archaeology with \hermes\ 
\citep[\galah;][]{DeSilva_2015} survey seeks to derive $\approx$30 precise 
abundances for $\sim10^6$ stars. 
Future ground-based surveys will derive abundances from $\approx$20 million 
spectra in each hemisphere \citep[e.g., \weave, \fourmost;][]{weave,4most},
and the radial velocity spectrometer on \gaia\ can deliver a few abundances 
for $\approx$150 million stars.

Chemical tagging is a simple idea, but with this volume of data it represents
a formidable parameter estimation and model selection problem.
Before reviewing the literature on chemical tagging, here we explicitly
formulate the scope of the analysis problem.

\subsection{What do we want to infer from chemical abundances?}
\label{section:the-problem}

Let us assume that a set of stars have been observed spectroscopically, and
chemical abundances have been derived from those data.
We will unrealistically assume that the data are noiseless, and the derived
abundances are unbiased, known with infinite precision, and complete (e.g., no 
missing abundances).
Regardless of the number of stars observed or abundances derived, there are a
number of inferences we seek to make:

\begin{enumerate}
  \item The number of star clusters (or star-forming events) where at least 
        one star was observed and chemical abundances were derived.
  \item The conditional probabilities (or memberships) of each star belonging
        to every star cluster or star-forming event.
        Alternatively, the relative weighting of individual mixtures.
  \item The mean abundances of each star cluster. 
  \item The covariance matrix of abundances of each star cluster.
        This covariance matrix includes the intrinsic (cluster) variance
        of individual abundances (e.g., the homogeniety), and allows for
        correlations between abundances.\footnote{
          It is likely -- or hoped -- that \emph{true} correlations between 
          abundances will be accounted for through the latent factors, and any 
          remaining correlations are the result of the measurements.}
  \item The \emph{number} of common multivariate latent factors, variables 
        that are not observable but affect all or many of the observations.
        In the context of chemical tagging, the yields of core-collapse
        supernovae can be considered a common multivariate latent factor: 
        those yields are not directly observable, but they contribute to much 
        of the data.
        Here we are interested in the \emph{number} of common multivariate
        latent factors. For example, the number of sources of enrichment:
        different classes of supernovae, asymptotic giant branch (AGB) stars, 
        etc.
  \item The multivariate factor loads of individual latent factors. 
        For example, if core-collapse supernovae and AGB stars had both 
        contributed to the chemical abundances of all stars in the sample, 
        then the \emph{factor loads} would be the \emph{yield} of the 
        supernovae, or the yield of relative abundances that were produced by 
        the AGB star. 
        Each star will be affected differently by those yields (see item 7).
  \item The relative \emph{scoring} of different latent factors on each star.
        This is a scaling that is applied to the latent yields. Continuing our
        analogy on yields from core-collapse supernovae and AGB stars, the
        \emph{factor scores} can be thought of as the relative contributions 
        of the different factor loads, for a single star.
  \item The intrinsic \emph{specific variances} of the individual abundances.
        These variances can be thought of as the intrinsic variability from 
        the different factor loads.
        The intrisic variances affect all observations, before accounting for
        any homogeneity in individual clusters.
  \item The \emph{number} of stellar age- or parameter-dependent latent 
        factors that affect the data. 
        The surface chemical abundances of stars \emph{do change} over its
        lifetime, at least in part due to atomic diffusion, veiling, and 
        thermohaling mixing.
        Understanding and modelling these effects will be critical for 
        chemical tagging stars across all stellar ages and evolutionary states.
  \item The factor loads of stellar age- or parameter-dependent latent factors.
        In this example, a single multivariate factor load may represent the
        age-dependent effects of atomic diffusion.
  \item The factor scores of stellar age- or parameter-dependent latent 
        factors.
\end{enumerate}

The number of things we seek to infer from the available chemical abundances
is vast.
Indeed, even if we ignore the latent factors then a non-trivial model
selection problem remains: What is the number of star clusters or star-forming 
events?
In the full description of chemical tagging we seek to know (1) the number of
star clusters, (2) the number of latent factors, and (3) the number of age-
or stellar parameter-dependent factors.
Going further, if we wanted to improve the accuracy and precision of our
solution then we may seek to include joint priors on the memberships of stars
being associated with specific clusters, either based on their ages or
astrometry.
In summary, simplified versions of chemical tagging represent formidable model
selection problems with non-convex\footnote{
  If a problem is convex then any local solution is mathematically guaranteed
  to be the global solution.
  If the problem is non-convex, then a local solution may not be the global
  solution.
} objective functions.

\subsection{Chemical tagging progress to date}
\label{sec:literature-review}

There has been substantial progress on chemical tagging, despite its difficulty.

% Most have opted to address individual components of what is listed above,
% in idealistic situations.

@@

\begin{comment}
\subsection{Problems using real world data}

There are a number of problems to consider when using real world data.

\begin{enumerate}
  \item There are missing data.
        Given high-resolution spectra with signal-to-noise (S/N) ratios of
        $\infinity$, there will be chemical abundances where only an upper
        limit can be reported.
        For example, if there are no transitions in the spectra, or the
        chemical abundance is so low that there is no net absorption in the 
        photosphere, then there will be no observable effect in the data.
        Chemical tagging approaches must be able to accept `upper limits'\footnote{
          The use of the term `\emph{upper limit}' amongst spectroscopists is
          muddled. 
          Frequently an `upper limit' is reported that represents (near-)zero
          likelihood that the abundance could be above the quoted value.
          Alternatively, a reported `upper limit' might describe a 
          $\approx1\sigma$ (non-symmetric) limit estimated from the local
          noise in the data.
          Ideally a posterior distribution should to be provided.
        }
        reported by spectroscopists.
  \item The abundance measurements may be biased.
  \item If there are uncertainties provided for abundances, then those
        uncertainties are likely to be underestimated.
  \item The measurements will have heteroscedastic noise properties.
        In other words, the abundance uncertainties will differ between
        elements and stars.
  \item The measurements of individual chemical abundances may be correlated
        due to the way the abundance was measured.
        For example, if there is a single absorption line present in the
        spectrum for an element (e.g., \ion{O}{1}), and that absorption line
        is blended with a comparably strong transition from another element
        (e.g., \ion{Ni}{1}), then the measured abundances of \ion{O}{1} and
        \ion{Ni}{2} will be (positively) correlated.
        This correlation may be difficult to discern from an \emph{intrinsic}
        correlation between the two elements.
\end{enumerate}
\end{comment}
